OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-02 main.tex
Wrote the methods section.

A common representation for the data yielded in a HiC experiment is a contact matrix, obtained slicing the genome in contiguous bins of fixed length (resolution) and computing the number of contacts between each pair of bins, corresponding to a cell in the contact matrix ($x_{ij}$), providing a quantification of the interaction between the tow loci.

Our proposal is based on the overall contacts registered for each bin, thus reducing the matrix to one dimension (hence the name). We assumed that the total number of contacts per bin ($t_{i}$) could be approximated by a negative binomial distribution with every potentially biasing genomic feature (known a priori) contributing independently to its mean ($\lambda_i$). Given that this relationship might not be linear (as can be seen in Figure \ref{fig:totals}A), we allowed a smooth representation using thin plate penalized regression splines \citep{wood2003thin} in a generalized additive model \citep{wood2011fast}. The model can be parametrized as follows:
$$
t_i = \sum_{i}{x_{ij}}
$$
$$
t_i \sim NB(\lambda_i, \theta) \\
$$
$$
log(\lambda_i) \propto \sum_{K}{f_k(z_k)}
$$
where $x_{ij}$ are the number of contacts between bins $i$ and $j$, and there is an additive term for each $z_k$ genomic feature. The smooth functions $\{f_k(z_k)\}$ are estimated jointly with the negative binomial dispersion parameter $\theta$ using R \citep{coreteam2014r} \texttt{mgcv} package \citep{wood2011fast}.%should we parametrize in terms of gamma-poisson instead negative binomial???

Once the model of the total counts has been estimated, we can further rescale  $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that can be used to compute the corrected counts ($c_{ij}$) or, alternatively, it can be included in downstream analysis:
$$
\lambda_i' = \frac{\lambda_i}{\sum_j^n{\lambda_j}/n}
$$
$$
c_{ij} = \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
$$

Following  previous works \citep{yaffe2011probabilistic,hu2012hicnorm} we included mappability, CG content and number of restriction sites (comparable to effective length) as genomic features, but the model and implementation allow easy extension with any available genomic feature.

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
	\caption{
		A Non linear relationship between a genomic feature (number of restriction sites per bin) and the total number of contacts per bin for a T47D sample. B Total number of contacts per bin in chromosome 17 of a T47D sample. Raw (in brown, noisier) and corrected (in blue, more precise) data are depicted.
	}\label{fig:totals}
\end{figure}

@@

As an environment to test the bias correction, we gather a set of published \citep{ledily2014distinct,encode2012integrated,rao20143d,stadhouders2017transcription,lin2012global,dixon2012topological} and unpublished data from HiC experiments of different cell types and organisms. The details can be found in Table \ref{tab:samples}. Briefly, we used several experiments comprising T47D breast cancer cell lines (7 samples), K562 leukemia cell lines (4 samples), both with aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples) cells, both with normal diploid karyotypes. The experiments used different protocols (\textit{i.e.} original HiC \citep{lieberman2009comprehensive} and in-situ HiC \citep{rao20143d}), different restriction enzymes (DpnII, HindIII, MboI and NcoI) and they were produced in distinct laboratories.

We also make use of array-based copy-number segmentation of the two cell lines obtained from COSMIC database \citep{forbes2010cosmic}.

@@

All data were processed though an in-house pipeline based on TADBit \citep{serra2016structural}. Briefly, after fastq quality control, paired reads were mapped against the corresponding reference genome (hg38 and mm10) taking into account the restriction enzyme motifs. Non-informative contacts were removed applying the following TADBit \citep{serra2016structural} filters: Self-circle, dangling-ends, error, extra dangling-end, duplicated and random-breaks. For more details, see methods of \cite{stadhouders2017transcription}. In addition, the pipeline is available, with minor modifications, in \cite{quilez2017managing} material.

We developed the routines contained in the \texttt{dryhic} R package to efficiently create sparse representations of contact matrices and further apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections. \texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp} \citep{wu2016computational} were used to get \textit{lgf} and \textit{caib} corrections respectively. All results are based on 100 Kbp resolution, although no major differences were found using different ones (data not shown).

\subsection{Total number of contacts as copy number proxy}

In euploid samples, we expect the coverage of a HiC experiment to be directly related to the number of copies (a duplicated region will show, on average, twice the number of contacts). An ilustrative example can be seen in Figure \ref{fig:copy_number}B. Given that our model is based on total number of contacts, which in turn is directly related to the number of copies, we reasoned that a preliminary test would be to check if the corrected number of contacts per bin ($ct_i = \sum_{i}{c_{ij}}$) reflects the copy number (as measured by an independent technique such as array-based copy-number segmentation) better than the alternatives.

\subsection{Sample comparison}

In order to test performance, we relied on pair-wise comparisons of contact matrices. However, there is little consensus on how to compare two HiC matrices. The simplest metric is the Spearman correlation applied to cis (intra-chromosomal) contacts up to a given distance (\textit{i.e.} 5 Mb). Though, the decay in the number of contacts with increasing genomic distance is the main driver of contact abundance and it could mask real differences between samples. This distance decay can be taken into account, for instance, computing the expected number of interactions given the genomic distance between any pair of loci. Thus, we can compare matrices in terms of observed over expected contacts, via the Pearson correlation (again focusing on a given distance range). Alternatively, we can compute the correlation per distance stratum and then obtain a stratum-adjusted correlation coefficient (SCC) as defined in \cite{yang2017hicrep}. Additionally, \cite{yan2017hicspector} proposed a reproducibility score based on the correlation between a set of the Laplacian matrix's eigenvectors (corresponding to the smallest eigenvalues) of the two experiments.

We defined three sets of samples: One with all T47D experiments plus two K562 samples, another with all K562 experiments plus two T47D samples and a third one comprising all mm10 samples (see Table \ref{tab:samples} for details). Given a set of experiments and a metric, we first computed all pair-wise combinations between experiments and then classified the comparisons according to the characteristics of each pair (\textit{i.e.} same / different cell types, same / different protocols, etc ...).

To benchmark our method (\textbf{\textit{oned}}), we also test the performance of other alternatives. Three of them were based on the same implicit iterative correction \citep{imakaev2012iterative}: A first alternative computing one iteration (\textbf{\textit{vanilla}}), a second one iterating until convergence (\textbf{\textit{ice}}) and a third one using the chromosome adjusted iterative correction (\textbf{\textit{caib}}) \citep{wu2016computational}. The fourth one was inspired on the explicit bias correction originally presented by \cite{yaffe2011probabilistic}, but based on the local genomic features (\textbf{\textit{lgf}}) approach introduced by \cite{hu2012hicnorm} as implemented in \cite{servant2012hitc}.

Besides, we compared the raw matrices and used these comparisons as a baseline to present the gain / loss of correlation relative to the uncorrected data, thus allowing us to verify the effect of the bias correction procedure. The differences relative to the raw comparisons were estimated using a linear mixed model with the method as fixed effect and the chromosome as a random effect, and using the \texttt{lmer} function of the \texttt{lme4} R package \citep{bates2015lme4}.

In an attempt to further summarize the results, we reasoned that the correlation between unrelated samples (\textit{i.e} different cell types) should be lower than the correlation between related samples (\textit{i.e} same cell types), so we could re-interpret each correlation metric as a classifier score. Thus, we computed the receiver operating characteristic (ROC) curves and corresponding area under the curve (AUC) using the R \texttt{ROCR} package \citep{sing2005rocr}.

@@

@@

@@

\begin{figure}
	\centerline{\includegraphics[width=.45\textwidth]{img/copy_number_figure2.pdf}}
	\caption{
		Pearson correlation between total number of contacts per bin and independent copy number estimation (COSMIC) for each of the methods compared. Left panel T47D breast cancer cell line, right panel K562 leukemia cell line. The new proposal (in blue) outperforms the rest of alternatives.
	}\label{fig:copy_number}
\end{figure}

@@

>>>>

The most common representation of Hi-C data is a contact matrix, obtained
by slicing the genome in $n$ consecutive bins of fixed size (the
resolution) and computing the number of contacts between each pair of
bins. The values are stored in the cells of the contact matrix ($x_{ij}$),
quantifying of the interaction between the tow loci at positions $i$ and
$j$.

Our approach is to model the tally of contacts for each bin, thus reducing
the matrix to a one dimension score (hence the name OneD). We assume that
the total number of contacts per bin ($t_{i}$) can be approximated by a
negative binomial distribution. This choice is sensible because the amount
of contacts is a discrete variable and because the negative binomial
distribution allows for overdispersion. We further assume that the
explicit sources of bias have independent contributions to the mean of the
distribution for a given bin ($\lambda_i$).

Given that this relationship might not be linear (see for instance
Figure~\ref{fig:totals}A), we allowed a smooth representation
using thin plate penalized regression splines \citep{wood2003thin} in a
generalized additive model \citep{wood2011fast}. The model can be
parametrized as

\begin{align*}
t_i &= \sum_j^n{x_{ij}} \sim  NB(\lambda_i, \theta), \\
\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)},
\end{align*}

\noindent
where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
$z_k$ is the additive bias of genomic feature $k$. The smooth functions
$\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
dispersion parameter $\theta$ using the \texttt{mgcv} package
\citep{wood2011fast} of the R software \citep{coreteam2014r}.

Once the parameters of the model are estimated, we rescale the estimated
means $\{\lambda_i\}$ to obtain a correction vector $\{\lambda_i'\}$ that
can be used to compute the corrected counts ($\hat{x}_{ij}$).

\begin{align*}
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}.
\end{align*}

In line with previous methods
\citep{yaffe2011probabilistic,hu2012hicnorm}, the features
used to fit the model are the local G+C content, read mappability and the 
content and number of restriction sites. The model and the implementation
can be easily extended with any user-provided genomic feature.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy, explaining that
the signal is approximately half compared to the rest of the chromosome.
Also notice that the raw signal is noisier than the corrected one.}
\label{fig:totals}
\end{figure}

@@

To test the correction of biases, we gathered a set of
published \citep{ledily2014distinct, encode2012integrated, rao20143d,
stadhouders2017transcription, lin2012global, dixon2012topological} and
unpublished Hi-C data of different cell types and organisms. The details
can be found in Table \ref{tab:samples}.

We used several experiments comprising T47D breast cancer cell
lines (7 samples), K562 leukemia cell lines (4 samples), both with
aberrant karyotypes, and mouse primary B (6 samples ) and ES (7 samples)
cells, both with normal diploid karyotypes. The experiments were carried
out in different laboratories, following either the original Hi-C protocol
\citep{lieberman2009comprehensive} or the newer \textit{in situ} version
\citep{rao20143d}, and using different restriction enzymes (DpnII,
HindIII, MboI and NcoI).

We also used array-based copy-number segmentation of the two cell
lines obtained from COSMIC database \citep{forbes2010cosmic}.

@@

All data were processed though a pipeline based on TADBit
\citep{serra2016structural}. Briefly, after controlling the quality of
fastq files, paired-end reads were mapped in the corresponding reference
genome (hg38 and mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADBit
filters: self-circle, dangling-ends, error, extra dangling-end, duplicated
and random-breaks. For more details, see the methods section of
\cite{stadhouders2017transcription}. In addition, the pipeline is
available from the supplementary material published by
\cite{quilez2017managing}.

We developed the routines contained in the \texttt{oned} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ice} and \textit{oned} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{lgf} and
\textit{caib} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences for different
values (not shown).

\subsection{Comparison of Hi-C matrices}

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mb in what follows). The second is
to compare the observed over expected contacts via the Pearson correlation
up to a given distance range. Compared to the first, this metric gives
more weight to changes occuring away from the diagonal. The third option
is to compute a correlation per distance stratum and then obtain a
stratum-adjusted correlation coefficient (SCC) as defined in
\cite{yang2017hicrep}.  Finally, \cite{yan2017hicspector} proposed a
reproducibility score based on the correlation between the eigenvectors of
the Laplacian matrix associated with the smallest eigenvalues. This
approach borrows the concepts of spectral clustering
\citep{von2007tutorial} and amounts to comparing high level features of
the matrix.

We defined three data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples (see Table
\ref{tab:samples} for details). Given a set of experiments and a metric,
we first computed all pairwise combinations between experiments and then
classified the comparisons according to the characteristics of each pair
(cell type, protocol, batch, and treatment).

We benchmark \textbf{\textit{oned}}, against methods
\textbf{\textit{vanilla}}, \textbf{\textit{ice}}
\citep{imakaev2012iterative}, \textbf{\textit{caib}}
\citep{wu2016computational} and \textbf{\textit{lgf}}
\citep{hu2012hicnorm, servant2012hitc}. The first three methods correct
biases implicitly, whereas the fourth method does it explicitly.

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were then
computed using \texttt{ROCR} package \citep{sing2005rocr}.

@@

%% --------------- Results --------------- %%

@@

In aneuploid samples, the local coverage of the Hi-C experiment is
proportional to the number of copies of the region (a duplicated region in
a diploid genome will have on average a 50\% increase in the number of
contacts). See Figure \ref{fig:copy_number}B for a typical example.
Given that our model is based on the total number of contacts, which in
turn is related to the number of copies, we reasoned that a
preliminary test would be to check if the corrected number of contacts per
bin ($\hat{t}_i = \sum_{i}{\hat{x}_{ij}}$) reflects the copy number (as
measured by an independent technique such as array-based copy-number
segmentation) better than the alternatives.

@@

\begin{figure}
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
	\caption{
    Pearson correlation between total number of contacts per bin and
independent copy number estimation (COSMIC) for each of the methods
compared. Left panel T47D breast cancer cell line, right panel K562
leukemia cell line. The new proposal (in blue) outperforms the rest of
alternatives.  }\label{fig:copy_number}
\end{figure}

@@

**
In an attempt to further summarize the results, we reasoned that the
correlation between unrelated samples (\textit{i.e} different cell types)
should be lower than the correlation between related samples (\textit{i.e}
same cell types), so we could re-interpret each correlation metric as a
classifier score.
**
